---
title: "In Vivo Statistical Analysis Report"
author: "Crown Bioscience"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
    number_sections: true
    fig_caption: yes
    keep_tex: true
    includes:
      in_header: header.tex
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"  
params:
  source: NA
  tv_file: NA
  tv_str: NA
  p_adj_method: BH	
  start_day: 0
  analysis_day: NA
  tgi_def: 3
  sel_group: NA
  ref_group: NA
  posthoc_cmps: all_pairwise
  analysis_var: TV
---

```{r setup, include=FALSE}
require(dplyr)
require(magrittr)
require(ggplot2)
require(reshape2)
require(RColorBrewer)
require(ggpubr)
require(multcomp)
require(PMCMRplus)
require(knitr)
require(tinytex)
library(kableExtra)
require(htmltools)
knitr::opts_chunk$set(echo = FALSE, fig.pos = "H", message=F, warning = F, error = TRUE)
options(knitr.table.format = "latex") 
```

```{r load}
grps_cols = c("gray0", "red","dark green","yellow","purple","blue","grey",
             "chartreuse1","light blue","palevioletred1","salmon4","peachpuff","gray48",
              "greenyellow","orange","salmon","purple2","lightgoldenrod1","brown4","darkorchid4")
mytheme <- theme(plot.title=element_text(face="bold.italic",size=16, color="brown",hjust=0.5),
                 axis.title.x=element_text(face="bold.italic",size=14, color="brown"),
                 axis.text=element_text(face="bold", size=10,color="darkblue"),
                 axis.title.y=element_text(face="bold",size=14, color="brown"),
                 strip.text = element_text(face="bold", size=12,color="brown"),
                 panel.background=element_rect(fill="white",color="darkblue"),
                 panel.grid.major.y=element_line(color="grey",linetype=1),
                 panel.grid.minor.y=element_line(color="grey",linetype=2),
                 panel.grid.minor.x=element_blank(),
                 legend.position="top")

vecSummary=function(v){
    mv=sprintf('%.2f',mean(v))
    n=length(v)
    sem=sprintf('%.2f',sd(v)/sqrt(n))
    paste0(mv,' +/- ',sem,'(',n,')')
}

#convert a matrix a vector
matrix2vec=function(m){
    vec=as.vector(m)
    names(vec)=as.vector(sapply(colnames(m),paste,rownames(m),sep=" - "))
    vec
}

#calculate TGI based on tumor volume data
tgi_defs=c('T/C','Delta(T)/Delta(C)','1-Delta(T)/Delta(C)')
getTGI=function(tv_long,start_d,ref_g,def=3){
    tvs=subset(tv_long,day>=start_d)
    mean_tv_gd=tapply(tvs$TV,list(tvs$Group,tvs$day),mean,na.rm=T)
    t_c=apply(mean_tv_gd,1,function(x) x/mean_tv_gd[1,])
    #print(mean_tv_gd)
    tv_diff=mean_tv_gd-mean_tv_gd[,1]
    tv_diff=t(tv_diff[,-1])
    dt_dc=tv_diff/tv_diff[,ref_g]
    tgi=1-dt_dc
    if(def==1) tgi=t_c
    if(def==2) tgi=dt_dc
    tgi=tgi[,colnames(tgi) != ref_g,drop=F]
    tgi=round(tgi,3)
}

p_adj_method=params$p_adj_method #use benjami-hochberg p-value adjust
start_day=as.numeric(params$start_day) #user specified start day, to calculate RTV
analysis_day=as.numeric(params$analysis_day) #user specified day for statistical analysis
tgi_def=as.numeric(params$tgi_def) #TGI definition
ref_group=params$ref_group #user specified reference group
sel_group=unlist(strsplit(params$sel_group,"\\s+"))
posthoc_cmps=params$posthoc_cmps
mcp_grps=NULL #user specified multi-group comparisons
analysis_var=params$analysis_var #TV or RTV, user specified variable for statistical analysis
source=params$source
tv=NA
if(source=='upload'){
   tv=read.csv(params$tv_file,header=T,stringsAsFactors = F) #read raw tumor volume data
}else if(source=='paste'){
   tv=read.csv(text=params$tv_str,header=T,stringsAsFactors = F)  
} 
#tv=read.csv('../../TV.csv',h=T)
#if(tv_str != '') tv=read.csv(text=tv_csv,header=T,stringsAsFactors = F) 
names(tv)[1:2]=c('Group','Mouse')
names(tv)=gsub('^X','Day',names(tv))
mouseTV=apply(tv[,c(-1,-2)],1,function(x) any(x>0)) #tv volume of a mouse >0 for any given day
tv=tv[mouseTV,]
tv$Group = gsub("^.*?(\\d+).*","G\\1",tv$Group)
tv$Group = as.factor(tv$Group)
n_grp = length(levels(tv$Group))
tv$Mouse = as.factor(tv$Mouse)
tv_long=melt(tv,id.vars=c(1,2),value.name="TV")
tv_long=subset(tv_long,!is.na(TV))
tv_long$day=as.numeric(substr(tv_long$variable,4,10))
fig_height=n_grp/2*5


#tv_long = subset(tv_long, is.numeric(TV) & TV>=0) #remove non-valid tv
#tv_long=subset(tv_long,day >= start_day)
tvn = with(tv_long,tapply(TV,list(day,Group),n_distinct))
dmo = apply(tvn,1,function(x) sum(x>=2,na.rm=T))
valid_days=as.numeric(names(dmo[dmo>=2]))
del_days=names(dmo[dmo<2])
del_days_str='* Statistical analysis requires at least two groups with two or more distinct observations.  
'
if(length(del_days)>0) del_days_str=paste0(del_days_str,'* Removing Day **',paste(del_days,collapse = ','),'** from further analysis  ')
#print(del_d)
#print(del_days_str)
tv_long = subset(tv_long,day %in% valid_days)
tv_long$logTV=log(tv_long$TV+1,10)

if(is.null(start_day)) start_day=min(tv_long$day)
tv_long=subset(tv_long,day >= start_day)
day0=ifelse(is.null(start_day),min(tv_long$day),start_day)
tv0=tv_long[tv_long$day==day0,c('Mouse','TV')]
names(tv0)[2]='TV0'
tv_long=merge(tv_long,tv0,by="Mouse")
tv_long$RTV=ifelse(tv_long$TV0==0,NA,tv_long$TV/tv_long$TV0)
tv_long$DeltaTV=tv_long$TV-tv_long$TV0
ndays=length(unique(tv_long$day))
fig_height_d=ndays/2*5
font_size_d=8
if(ndays>8) font_size_d=7
if(ndays>12) font_size_d=5
if(ndays>15) font_size_d=4
font_size_g=7
if(n_grp>10) font_size_g=6
#mean_tv=with(tv_long,tapply(TV,list(Group,day),mean,na.rm=T))
```

\newpage
#Raw tumor volume data
`r del_days_str`
```{r}
#print(c(sel_group,ref_group))
colors=brewer.pal(n_grp,'Set3')
#colors=rainbow(n_grp)
grps_nm=table(tv$Group)
rgs=c(0,cumsum(grps_nm))
k1=tv %>% kable(caption='Raw Tumor Volume data',row.names=F,longtable = T, booktabs = T,linesep = "",digits = 2) %>% 
    kable_styling(latex_options = c("striped", "scale_down","repeat_header","HOLD_position"),font_size = font_size_d,full_width=F) 
for(i in 1:n_grp){
    color_ind = i %% 12
    if(color_ind==0) color_ind=12
    k1 = row_spec(k1,(rgs[i]+1):rgs[i+1],background=colors[color_ind])
}
if(ndays>15) k1=landscape(k1)
k1
```

\newpage
#Tumor volume data summary
##Tumor volume summary statistics
```{r}
tv_summary=with(tv_long,tapply(TV,list(Group,day),vecSummary)) #A summary of tumor volume statistics,in (mean +/- sem(n) format)
tv_s=tv_summary %>% kable(caption='Tumor Volume Data summary, formatted as Mean +/- SEM(n)', linesep = "", booktabs = T) %>% 
    kable_styling(latex_options = c("striped","scale_down","HOLD_position"),full_width=F) %>%
    column_spec(1,width='8em',bold=T)
if(ndays>15) tv_s=landscape(tv_s)
tv_s
```


##Tumor growth curves
```{r fig1,fig.width=12,fig.height=fig_height,fig.cap="Tumor growth curves"}
#plot tumor growth curve for each group
ggplot(tv_long)+aes(day,TV,group=Mouse,color=Group)+scale_color_manual(values=grps_cols)+
    labs(title='Tumor growth curves',x='Day',y=expression(paste('Tumor Volume ( ',mm^3,')')))+
    geom_point()+geom_line()+facet_wrap(~Group,ncol=2)+mytheme+theme(legend.position = 'none')
```

##Tumor volume Mean+/-SEM plot
```{r fig2,fig.cap="Tumor volume Mean+/-SEM plot"}
ggline(tv_long,x='day',y='TV',add=c('mean_se'),group = 'Group',title='Tumor volume Mean+/-SEM plot',
       xlab='Day',ylab=expression(paste('Tumor Volume ( ',mm^3,')')),size=1.5,color='Group',
       add.params =   list(size=1.5,alpha=0.8,position='dodge'))+scale_color_manual(values=grps_cols)+mytheme
```

##Compare `r analysis_var` among groups by day, p-value calculated by Kruskal-Wallis test
```{r fig3,fig.width=12,fig.height=fig_height_d,fig.cap="Compare tumor volume among groups by day"}
p=ggboxplot(tv_long,x='Group',y=analysis_var,title=paste0(analysis_var,'~Group|Day Boxplot'),add=c('jitter'),color='Group',
          xlab='Day',facet.by = 'day')+scale_color_manual(values=grps_cols)+
    stat_compare_means(aes(label = paste0("p=", ..p.format..)),label.x.npc = 'left',label.y.npc = 'top')+mytheme
facet(p,facet.by = 'day',scales='free',ncol=2)
```

\newpage
#Nominal p-values derived from t-test or Mann-Whitney U test
```{r pval}
twoSampleTest=function(v1,v2){
    v1=v1[!is.na(v1)]
    v2=v2[!is.na(v2)]
    if(n_distinct(v1)<2 | n_distinct(v2)<2) return(NA)
    pb=bartlett.test(list(v1,v2))$p.val #bartlett test 
    pval=NA
    if(pb<0.05){
        #pval=round(wilcox.test(v1,v2)$p.val,3)
        pval=sprintf("%.3g",wilcox.test(v1,v2)$p.val)
        #pval=wilcox.test(v1,v2)$p.val
        pval=cell_spec(pval,color=ifelse(as.numeric(pval)<0.05,'red','blue'),
                       bold = ifelse(as.numeric(pval)<0.01,T,F),background = 'lightgray')
    }else{
        pval=sprintf("%.3g",t.test(v1,v2)$p.val)
        #pval=t.test(v1,v2)$p.val
        #pval=round(t.test(v1,v2)$p.val,3)
        pval=cell_spec(pval,color=ifelse(as.numeric(pval)<0.05,'red','blue'),
                       bold = ifelse(as.numeric(pval)<0.01,T,F))
    }
    return(pval)
}
#pairwise comparison between groups for some specific day
cmpDay=function(df,var_sel){
    groups=as.character(levels(df$Group))
    gcs=combn(groups,2)
    pvals=apply(gcs,2,function(x) twoSampleTest(df[df$Group==x[1],var_sel],df[df$Group==x[2],var_sel]))
    names(pvals)=apply(gcs,2,function(x) paste0(x[1],' vs ',x[2]))
    pvals
}

days=split(tv_long,tv_long$day)
nday=length(days)
pval_table=do.call(rbind,lapply(days,cmpDay,analysis_var)) #generate norminal p-value table, p-values from mann-whitney test are ended with *
#pval_table=pval_table[order(rownames(pval_table)),]
#pairwise.t.test(days[[1]]$TV,days[[1]]$Group,pool.sd = F,p.adjust.method = 'none')
longtable_flag=ifelse(ncol(pval_table)>30,TRUE,FALSE)
pval_t= t(pval_table) %>% kable(escape=F,linesep = "", booktabs = T,longtable=longtable_flag,digits = 2,caption=paste0('Comparing two independent groups ',analysis_var,': Nominal p-values')) %>% 
    kable_styling(latex_options = c("striped","scale_down","HOLD_position"),full_width=T,font_size = font_size_d) %>%
    add_header_above(c("","Day"=nday),bold = T) %>%
    column_spec(1,bold=T,border_right = T) %>%
    footnote(general='p-values from Mann-Whitney U test are denoted with a lightgray background, and p-values smaller than 0.05 are colored red')
if(ndays>15) pval_t=landscape(pval_t)
pval_t
```

\newpage
#TGI calculation for first day=`r start_day` and `r ref_group` as vehicle group
```{r}
tgi_table=getTGI(tv_long,start_day,ref_group,tgi_def)
tgi_table %>% kable(escape=F,linesep = "", booktabs = T,caption='TGI table') %>%
    kable_styling(latex_options = c("striped", "scale_down","HOLD_position"),font_size = font_size_g,full_width=T) %>%
    add_header_above(c("","Group"=ncol(tgi_table)),bold = T) %>%
    column_spec(1,bold=T,border_right = T) %>%
    footnote(general=paste0('TGI defined as ',tgi_defs[tgi_def],', calculation based on first day=',start_day,' and ',ref_group,' as vehicle group'))
```

\newpage
#Statistical analysis for `r analysis_var` at Day `r analysis_day`

##Compare if groups are equal without multiple comparison correction for `r analysis_var` at Day `r analysis_day`
```{r fig4,fig.width=12,fig.asp=0.618,fig.cap="Compare Tumor Volume at Selected Date"}
pct_cutoff=0.8 #percentage cutoff; if analysis day is not specified, pick the last day with valid TV for over pct_cutoff in every group  
if(is.null(analysis_day)){
    grp_mouse_n=tapply(tv_long$Mouse,tv_long$Group,function(x) length(unique(x)))
    day_grp_n=table(tv_long$day,tv_long$Group)
    day_grp_pct=apply(day_grp_n,1,function(x) x/grp_mouse_n)
    analysis_day=max(as.numeric(colnames(day_grp_pct)[apply(day_grp_pct,2,function(x) all(x>=pct_cutoff))]))
}


tv_day=subset(tv_long,day==analysis_day & Group %in% sel_group)
tv_day$Group = factor(tv_day$Group)
tv_day$Group=relevel(tv_day$Group,ref=ref_group)
grps=sel_group
#run bartlett test to test homogeniety of variance and normality
pb_day=bartlett.test(tv_day$TV,tv_day$Group)$p.val
all_cmp=as.list(as.data.frame(combn(grps,2)))
all_cmp=lapply(all_cmp,as.character)
sel_cmp=rep(T,length(all_cmp))
if(posthoc_cmps == 'treatment_vs_vehicle') sel_cmp = sapply(all_cmp,function(x) ref_group %in% x)
test=ifelse(pb_day<0.05,'wilcox.test','t.test')
p = ggbarplot(tv_day, x = "Group", y = analysis_var, color='Group',fill=NULL,size=2,
              add=c('mean_se','jitter'),add.params = list(size=1.5),
              title=paste0('Day ',analysis_day,'-All pairwise comparison: ', test,' nominal p value'))
#p = ggboxplot(tv_day, x = "Group", y = "TV", add='jitter')
p = p+scale_color_manual(values=grps_cols)+stat_compare_means(method=test,comparisons = all_cmp[sel_cmp],label='p.signif')+mytheme+theme(legend.position = 'none')
print(p)
```

##Compare if groups are equal with multiple comparison correction for `r analysis_var` at Day `r analysis_day`

1. Test of homogeneity of variance by **Bartlett's test**
3. If p-value of Bartlett's test is **equal to or greater than 0.05**:  
    + When comparing two groups, run **Welch's t-test** to test if the means of two groups are equal
    + When comparing more than two groups:
        + Run **ANOVA** to test if the means of all groups are equal  
        + Run **Tukey HSD** for all pairwise comparisons or **Dunnett's t test** for each treatment against one control 
4. If p-value of Bartlett's test is **smaller than 0.05**:
    + When comparing two groups, run **Mann-Whitney U test** to test if the medians of two groups are equal
    + When comparing more than two groups:
        + Run **Kruskal-Wallis test** to test if the medians of all groups are equal  
        + Run **Conover's non-parametric all-pairs comparison test** for all pairwise comparisons, or **Conover's non-parametric many-to-one comparison test** for each treatment against one control 
    
```{r}
##if number of group is two, then do t-test or mann-whitney U test; or else run ANOVA/kruskal-wallis test and subsequent 
##posthoc analysis
grp_n=length(grps)
rps=NULL #p-values of two group comprison
omnibus_p=NULL #p-value of omni-bus test (ANOVA/Kruskal-wallis test)
posthoc_test=''
if(grp_n==1){
  print('Only one group,No statistical analysis will be done')      
}else if(grp_n==2){
  rps=ifelse(pb_day<0.05,wilcox.test(tv_day[,analysis_var]~tv_day$Group)$p.val,
             t.test(tv_day[,analysis_var]~tv_day$Group)$p.val)
  omnibus_p=rps
  names(rps)=ifelse(pb_day<0.05,'Mann-Whitney U test','T test')
  names(omnibus_p)=ifelse(pb_day<0.05,'Kruskal-Wallis test','One-way ANOVA')
}else{
  if(pb_day<0.05){
      omnibus_p = kruskal.test(tv_day[,analysis_var]~tv_day$Group)$p.val
      names(omnibus_p)='Kruskal-Wallis test'
      posthoc_test = ifelse(posthoc_cmps=='treatment_vs_vehicle','kwManyOneConoverTest','kwAllPairsConoverTest')
  }else{
      aov_m=aov(tv_day[,analysis_var]~tv_day$Group)
      omnibus_p = summary(aov_m)[[1]]$Pr[1]
      names(omnibus_p)='One-way ANOVA'
      posthoc_test = ifelse(posthoc_cmps == 'treatment_vs_vehicle','dunnettTest','tukeyTest')
  }
  pvals=do.call(posthoc_test,list(tv_day[,analysis_var],tv_day$Group))$p.val
  rps=matrix2vec(pvals)
  rps=rps[!is.na(rps)]
}

all_ps=c(pb_day,omnibus_p,rps)
test_names=c("Bartlett's test",names(omnibus_p),names(rps))
   

#significance category, <0.05 *, <0.01 **, <0.001 ***
sigs=rep('ns',length(all_ps))
sigs[all_ps<0.05]='*'
sigs[all_ps<0.01]='**'
sigs[all_ps<0.001]='***'
output_df=data.frame('Test'=test_names,'p values'=sprintf("%.3g",all_ps),'significance level'=sigs,stringsAsFactors = F)
rownames(output_df)=NULL
output_df$p.values=as.numeric(output_df$p.values)
#print(output_df)
if(grp_n==2) output_df=output_df[-2,]
out_table = output_df %>%  mutate(p.values=cell_spec(p.values,color=ifelse(p.values<0.05,'red','blue'),bold = ifelse(p.values<0.01,T,F))) %>%
    kable(digits=3,escape=F,booktabs = T,caption=paste0('Comparing two independent groups ',analysis_var, ' at day ',analysis_day)) %>% 
    kable_styling(latex_options = c("striped", "scale_down","HOLD_position"),full_width=T) %>%
    column_spec(1,bold=T,border_right = T) %>%
    group_rows("Test of homogenity of variance and normality", 1, 1, bold=T, hline_after = T)
if(grp_n>2){
    out_table %>% group_rows("Test of overall equality among groups", 2, 2, bold=T, hline_after = T) %>%
    group_rows("Test of equality between individual groups", 3,nrow(output_df), bold=T, hline_after = T) 
}else{
    out_table %>% group_rows("Test of equality between two groups", 2,nrow(output_df), bold=T, hline_after = T) 
}    
```

```{r, results='asis'}
output_str=paste0('##Conclusions
- There are a total of ',grp_n,' groups. ',ref_group, ' is used as the control group.
')
if(pb_day<0.05){ 
    output_str=paste0(output_str,"- Because p-value of Bartlett's test is ",round(pb_day,2)," the homogeneity of variance assumption is violated, we run non-parametric test to compare groups.
  ")
}else{
    output_str=paste0(output_str,"- Because p-value of Bartlett's test is ",round(pb_day,2),", we run parametric test to compare groups.
  ")
}
if(grp_n==2){
    if(rps<0.05){
    output_str=paste0(output_str,"- Because p-value is ",sprintf("%.3g",rps), ", **two groups are not equal**.")
    }else{
        output_str=paste0(output_str,"- Because p-value of is  ",sprintf("%.3g",rps), ", **no significant difference is detected between two groups**.")
    }
}else if(grp_n > 2){
    if(omnibus_p<0.05){
    output_str=paste0(output_str,"- Because p-value of ", names(omnibus_p), " is ",sprintf("%.3g",omnibus_p), ", **Group comparisons with red font p-values are not equal**.")
    }else{
        output_str=paste0(output_str,"- Because p-value of ", names(omnibus_p), " is ",sprintf("%.3g",omnibus_p), ", **no significant difference is detected among all groups**.")
    }
}    
cat(output_str)
```


